---
title: "Networks in Psychology Project"
author: "Bhuvanesh Wadhwani"
date: "2023-02-16"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```



# Loading necessary libraries
```{r libraries}

packages_to_use<- c("tidyverse", "igraph", "igraphdata", "qgraph", "influenceR") #packages to use. if installed, load. if not installed, install then load. 

for(i in packages_to_use){
  if( ! i %in% rownames(installed.packages())  ) {
    print(paste(i, "not installed; installing now:\n") )
    install.packages(i)
  }
  
  require(i, character.only = TRUE)
}

```




# Take in data
```{r}

gossip.dat <- read.csv("data_final.csv", header = TRUE, row.names = 1) #loading gossip file that contains negative and positive gossip data

```

# Negative Gossip
```{r negative gossip}

ng.dat <- gossip.dat %>%
  select(NegGosF1R:NegGosM21R) #selecting columns for negative gossip from file

head(ng.dat) #check if the correct columns were selected

ng.mat <- as.matrix(ng.dat) #convert to matrix
ng.am <- graph_from_adjacency_matrix(ng.mat, mode = "undirected") #convert into igraph object and make sure it is undirected. It should be unweighted by default, so no code needed to run that. 

ng.am <- simplify(ng.am) #making sure there is only 1 edge between each pair of nodes to remove instances of A -> B and B -> A counting as 2 edges rather than 1. 
is.simple(ng.am) #check simplify function

summary(ng.am) #summary of igraph object
#44 nodes, 264 edges
```

# Visualization Negative Gossip
```{r, fig.height = 8, fig.width = 8}
par(mar=c(0,0,0,0)+.1)  #reduce borders
set.seed(111) #set seed for the same graph layout for reproducability

#plot negative gossip for visualization
plot(ng.am, layout = layout_with_graphopt(ng.am), #layout to view the graph
     vertex.frame.color = 'black', 
     vertex.label.family = 'sans', #readable font
     vertex.size = 8, 
     vertex.color = 'white', 
     vertex.label.cex = 0.7)

```

# Positive Gossip
```{r positive gossip}

pg.dat <- gossip.dat %>%
  select(PosGosF1R:PosGosM21R) #selecting columns for positive gossip. 

head(pg.dat) #check if the correct columns were selected

pg.mat <- as.matrix(pg.dat) #convert to matrix
pg.am <- graph_from_adjacency_matrix(pg.mat, mode = "undirected") #convert into igraph object and make sure it is undirected. It should be unweighted by default, so no code needed to run that.

pg.am <- simplify(pg.am) #making sure there is only 1 edge between each pair of nodes to remove instances of A -> B and B -> A counting as 2 edges rather than 1.
is.simple(pg.am) #check simplify function

summary(pg.am) #summary of igraph object
#44 nodes, 475 edges
```

# Visualization Positive Gossip
```{r, fig.height = 8, fig.width = 8}
par(mar=c(0,0,0,0)+.1) #reduce borders
set.seed(111) #set seed for reproducability 

plot(pg.am, layout = layout_with_graphopt(pg.am), 
     vertex.frame.color = 'black',
     vertex.label.family = 'sans',
     vertex.size = 8, 
     vertex.color = 'white', 
     vertex.label.cex = 0.7)

```

# Key Player Analyses
```{r key player analyses}
set.seed(111) #set seed for reproducability  
keyplayer_ng <- keyplayer(ng.am, k = 4) #KPP-Pos. selecting 4 key players from 44 nodes from negative gossip network. 4 was chosen as an estimated number. 
print(keyplayer_ng) #show the 4 key players negative gossip network
#F12, F23, M3, M12


set.seed(111) #set seed for reproducability
keyplayer_pg <- keyplayer(pg.am, k = 4) #KPP-Pos. same as above, but for positive gossip network. 4 was chosen again as an estimated number
print(keyplayer_pg) #show the 4 key players in positive gossip network
#F12, F16, F17, M2

```


# visualization negative gossip key player analysis
```{r, fig.height = 10, fig.width = 10}
# color nodes with highest scores 
V(ng.am)$color <- 'skyblue' #node color
V(ng.am)[keyplayer_ng]$color <- 'limegreen' #select different color for key players to view in the graph 

par(mar=c(0,0,0,0)+.1) #reduce borders

plot(ng.am, layout = layout_with_graphopt(ng.am), 
     vertex.frame.color = 'black', 
     vertex.label.family = 'sans',
     vertex.size = 8, 
     vertex.color = V(ng.am)$color, 
     vertex.label.cex = 0.8,
     vertex.label.color = 'black')


```


# Visualization positive gossip key player analysis
```{r, fig.height = 10, fig.width = 10}
# color nodes with highest scores 
V(pg.am)$color <- 'skyblue' #node color
V(pg.am)[keyplayer_ng]$color <- 'limegreen' #select different colors for key players to view in the graph  

par(mar=c(0,0,0,0)+.1) #reduce borders

plot(pg.am, layout = layout_with_graphopt(pg.am), 
     vertex.frame.color = 'black', 
     vertex.label.family = 'sans',
     vertex.size = 7, 
     vertex.color = V(pg.am)$color, 
     vertex.label.cex = 0.8,
     vertex.label.color = 'black')


```



# Effective Network Size (ENS) Negative Gossip
```{r ENS ng}
ens_data_ng <- data.frame(node = V(ng.am)$name,
                            ens_score_ng = ens(ng.am)) #ens scores for negative gossip. Getting ens scores (code: ens(ng.am)) with node name (code: V(ng.am)$name), and putting it into a data frame. 

# top 4 ens nodes 
arrange(ens_data_ng, -ens_score_ng) %>% head(4) #filter in descending order (code: -ens_score_ng) and show the top 4 nodes with the highest ens scores (code: head(4)) 
#M12, F12, F21, M3

```


# Visualization Negative Gossip ENS
```{r, fig.height = 8, fig.width = 8}
# color nodes with highest ENSes 
V(ng.am)$color <- 'skyblue'
V(ng.am)[name %in% c('NegGosM12R', 'NegGosF12R', 'NegGosF21R', 'NegGosM3R')]$color <- 'green' #highest ens nodes color

par(mar=c(0,0,0,0)+.1)  #reduce borders

plot(ng.am, 
     layout = layout_with_graphopt(ng.am), 
     vertex.frame.color = 'black', 
     vertex.label.family = 'sans',
     vertex.size = 8, 
     vertex.color = V(ng.am)$color, 
     vertex.label.cex = 0.6,
     vertex.label.color = 'black')
```


# Effective Network Size (ENS) Positive Gossip
```{r ENS pg}
ens_data_pg <- data.frame(node = V(pg.am)$name,
                            ens_score_pg = ens(pg.am)) #ens scores for positive gossip. Getting ens scores (code: ens(pg.am)) with node name (code: V(pg.am)$name), and putting it into a data frame.

# top 4 ens nodes
arrange(ens_data_pg, -ens_score_pg) %>% head(4) #filter in descending order (code: -ens_score_pg) and show the top 4 nodes with the highest ens scores (code: head(4))

#F2, F12, M12, F7

```

# Visualization Positive Gossip ENS
```{r, fig.height = 8, fig.width = 8}
# color nodes with highest ENSes 
V(pg.am)$color <- 'skyblue'
V(pg.am)[name %in% c('PosGosF2R', 'PosGosF12R', 'PosGosM12R', 'PosGosF7R')]$color <- 'green' #highest ens nodes color

par(mar=c(0,0,0,0)+.1) 

plot(pg.am, 
     layout = layout_with_graphopt(pg.am), 
     vertex.frame.color = 'black', 
     vertex.label.family = 'sans',
     vertex.size = 8, 
     vertex.color = V(pg.am)$color, 
     vertex.label.cex = 0.6,
     vertex.label.color = 'black')
```




# Centrality measures for negative gossip network
```{r}

degree(ng.am) |> sort(decreasing = T) |> head(4) #top 4 node degree in negative gossip network. Done by sorting in decreasing order (code: sort(decreasing = T) then showing top 4 scores (code: head(4)). 

closeness(ng.am)|> sort(decreasing = T) |> head(4) #top 4 node closeness in negative gossip network.Done by sorting in decreasing order (code: sort(decreasing = T) then showing top 4 scores (code: head(4)). 

#Top 4 for both degree and closeness = M12, F12, M5, F21. Both showed the same 4.

```

# centrality measures for positive gossip network
```{r}

degree(pg.am) |> sort(decreasing = T) |> head(4) #top 4 node degree in positive gossip network. Done by sorting in decreasing order (code: sort(decreasing = T) then showing top 4 scores (code: head(4)). 

closeness(pg.am)|> sort(decreasing = T) |> head(4) #top 4 node closeness in positive gossip network. Done by sorting in decreasing order (code: sort(decreasing = T) then showing top 4 scores (code: head(4)). 

#Top 4 for both degree and closeness = F2, F12, M12, M7. Both showed the same 4.

```

















